New seismological constraints on the available parameter space of macroscopic dark matter

### 2016-09-06 LaTeX/Article_Draft.tex
Update

The moon is an attractive site to carry out a seismological search for Macro impacts. Unlike the Earth, the Moon is nearly dormant. Most internal seismological activity in the Moon originates from deep Moonquakes, which are very attenuated by the time they reach the surface. Most of the noise in lunar seismograms comes from asteroids, but these are for the most part easily identifiable as such. Thus, in the case that we can distinguish between asteroids and Macros, the Moon is a nearly ideal location to place constraints via total seismic event rate. However, in the case that the Macro impacts are very weak, it may be impossible to distinguish them from asteroids or weak Moonquakes based on a single seismogram. In this case, it is necessary to identify correlated events among lunar seismometers which correspond to a propagating source of seismic waves. This sort of analysis will be left to future work.
\section{The Seismic Source}
\subsection{Seismic Wave Generation}
The energy from the initial impact generates a column of melted rock. The super-heated rock, held at constant volume, generates a pressure which sources the seismic waves that can be measured at detectors. Macro's with velocity insignificantly changed by their encounter with the Moon source an overpressure pressure of the form
\begin{align}
\begin{array}{cc}p = p_0 f(r)&\te{ if }a<r<b,-\f h2<z<\f h2\,,\\ p = 0&\te{ else,}\end{array}
\end{align}
where $f$ is an analytic function on $[a,b]$ whose range is a subset of $[-1,1]$, and $p_0$ is the maximum differential pressure in this region. After a short time, the pressure is still of this form, since the inhomogeneity of the Moon is significant only on large scales.

The seismometers on the moon have a net useful range of up to $16\,\te{Hz}$ and are sensitive to displacements as small as $0.3\,\text{nm}$ \cite{latham1973lunar}\cite{nakamura1982apollo}. Therefore, we only calculate the energy contained in the modes of frequency less than $20\,\te{Hz}$\cite{nakamura1982apollo}.

@@

\section{Seismic Wave Propagation}
The velocity of $p$-waves as a function of distance from the center $r$ of the Earth and Moon (and presumably other spherical rocky celestial bodies) is of the form $v = a^2 - b^2 r^2$ \cite{garcia2011very}\cite{dziewonski1981preliminary}. Using Snell's law, we obtain the differential equations for the trajectory of $p$-wave rays within such bodies
\begin{align}
v^2 \notag&= \dot r^2 + \f{p_\te{ray}^2 v^4}{r^2}\,,\\
0 &=\dot\theta\pm\f{p_\te{ray} v^2}{r^2}\,.
\end{align}
where $r$ is the radial coordinate of the ray, $\theta$ is the polar angle of the ray measured from the center of the body, and $p_\te{ray}$ is the ray parameter, which is fixed along a ray trajectory. These equations can be integrated. In the case that $v = a^2 - b^2 r^2$, we obtain for some constants $q$ and $\theta_0$.
\begin{align}
\notag&r(t) = \f{a\sqrt{(q^2e^{2abt}-b)^2 + 4a^2b^4 p_\te{ray}}}{b\sqrt{(qe^{2abt}+b)^2 + 4a^2b^4 p_\te{ray}}}\,,\\
&\tan(\theta(t) - \theta_0) = \f{qe^{4abt}-b^2}{4ab^3p_\te{ray}}+abp_\te{ray}\,.
\end{align}
\textit{A priori} these are geodesics on the Poincar\'e disc of radius $\f ab$, which is easily seen from the form of $v$.

@@

\section{Seismic Wave Detection}
Consider a $p$-wave traveling towards positive $x$ and of compact support $S$ in the plane normal to to its motion. Denote $A =\int_S\diff y\diff z$ The displacement field is
\begin{align}
u(\vec x,t) &= \chi_{\R\times S}\int\f{\diff k}{2\pi}U(k)e^{-\I k(x-v_p t)}\,,\\
\abs{u(\vec x,t)}&\leq\int\f{\diff k}{2\pi}\abs{U(k)}\,,
\end{align}
and the energy is
\begin{align}
E = \rho v_p^2 A\int\f{\diff k}{2\pi}k^2\abs{U(k)}^2\,.
\end{align}
As before, we denote the energy in the low frequency spectrum
\begin{align}
E_k = \rho v_p^2 A\int_0^k\f{\diff k'}{2\pi}k'^2\abs{U(k')}^2\,.
\end{align}
It follows that
\begin{align}
\abs{U(k)} = \sqrt{\f{2\pi}{\rho v_p^2 A}}\f1k\sqrt{\td{E_k}{k}{}}\,,
\end{align}
and from the estimate above
\begin{align}
\abs{u(\vec x,t)}\leq(2\pi\rho v_p^2 A)^{-1/2}\int_0^k\f{\diff k'}{k'}\sqrt{\td{E_{k'}}{k'}{}}\,.
\end{align}
In our case $\Xi E= C_1 k^2$ before attenuation. After anelastic attenuation, $E_k =\te{Ref}\,2C_1 k e^{-k\,\te{Att}}$
\begin{align}
\abs{u(\vec x,t)}\leq\sqrt{\f{2}{\rho v_p^2 A}}\sqrt{\f{\te{Ref}\,\Xi E}{\te{Att}}}\f1k\te{Erf}\ps{\sqrt{\f{\te{Att}\,k}{2}}}\,.
\end{align}
(25) is impossible to integrate when $E_k$ consists of the energies of two different rays, each with different attenuation factors. If attenuation were important, we may average $\te{Att}$ among the coincident rays and obtain an approximate expression. However, since we only wish to consider the lowest frequency modes, $f<20\te{Hz}$, the suppression $\te{Att} k$ is order unity, and does not contribute significantly to $u$. Taking $\te{Att} = 0$, the displacement caused by the incident $p$-wave is bounded above by
\begin{align}
\abs{u(\vec x,t)}&\leq\sqrt{\f{4\te{Ref}\,\Xi E}{\pi k\rho v_p^2 A}}\,,\\
&=\sqrt{\f{9\te{Ref}}{2mn}\f{\kappa\rho\ell}{\pi v_p^3A}}\f{\sqrt{ f_\te{max}}\sigma_Xv_X^2}{p_0}\,.
\end{align}
(28 will be deleted, I'm just using it to remember the formula)
\section{Simulation}
In previous works, e.g. \cite{banerdt2005seismic}, the homogeneous Earth and Moon models have been used to place constraints of the parameter space of macroscopic dark matter. This approach neglects lensing which occurs because of the velocity gradient and the spherical boundaries of strata, and neglects the energy losses from reflection and refraction across these boundaries. We use the analysis in section III to propagate rays, each carrying a fraction of the total energy deposited by the Macro, to the boundary of the Moon. We then use HEALPix to create an intensity map of the lunar surface for a representative sample of Macro impacts. We then convert the intensity maps to displacement maps, and using the sensitivity of the lunar seismometers, we obtain an average event rate that the lunar seismometers would measure.
\subsection{Data Generation}
We consider a Macro trajectory which passes a distance $D$ from the center of the moon. The Macro trajectory has length $L$, and along its trajectory we consider a set of $M$ points. From each of these points we propagate $N$ randomly oriented rays, each endowed with $E_i = \rho(r_{i})\sigma_Xv_X^2 L/(MN)$ energy, where $i\in\{1,\dots,M\}$ labels the point, and $\rho$ is taken to be a function of the radius. Because of the spherical symmetry, each ray is confined to a subspace of two dimensions. We use the trajectories in section III to propagate the rays. Since the moon is stratified, we must halt the propagation at the boundary of each layer to reflect and transmit rays. During a ray's propagation through a given stratum, its time spent in that layer $\Delta t_i$ is recorded, the attenuation factor in that layer $\te{Att}_i$ is recorded, and the reflection/refraction coefficient $\te{Ref}_i$ is recorded. Each time a ray makes contact with the surface of the Moon, its position, $\te{Att}$, $\te{Ref}$, and $\Delta t$ are recorded. The data produced by the propagation routine is a table of these values and the corresponding surface coordinates, as well as density at the source, $\kappa$ at the source, and $p$-wave velocity at the source. In this model, it takes 16 iterations to propagate a ray from one side of the moon to the other. Rays which propagate from one side of the moon to the other only after 16 iterations will be weak from undergoing 16 refractions. Thus in order to save on computation time, it is reasonable to reduce the number of iterations to 9.
\subsection{Data Analysis}
Analysis begins by converting the position data to HEALPix coordinates for a given $n_\te{side}$. We pick $n_\te{side} = 4$ since this corresponds to an angular resolution of 7 degrees, which accounts for the effects of diffraction about the strata boundaries. Define $A_\te{pix} = 4\pi R_\te{moon}^2/n_\te{pix}$ the area of a HEALPix pixel, where $n_\te{pix} = 12\times 2^{n_{\te{side}}}$. The time it takes a ray to cross a HEALPix pixel can be estimated $t_\te{crossing}^2 = A_\te{pix}/\pi v_\te{surface}^2$. According to VPREMOON, $v_\te{surface} = 1\,\te{km s}^{-1}$. Rays that contact a pixel within $t_\te{crossing}$ of one another are considered to add constructively. Since the integral in section IV is unreasonable to do for more than one ray, we average $\te{Att}$ over the coincident rays. We require that each ray distributes its energy evenly over the area of the HEALPix pixel, and thus set $A = A_\te{pix}$. This is a good approximation when $N=n_\te{pix}$, since the average angular separation of the rays at the source corresponds to the average angular separation of the HEALPix pixels.

@@

This procedure is repeated for 17 values of $D$ spaced evenly by $100\,\te{km}$. The average fraction of detectable surface area is then approximately
\begin{align}
\scr F =\f{\sum_D \scr F_D}{\sum_D D}\,.
\end{align}
If there are $n$ lunar seismometers, we suppose that each occupies its own HEALPix pixel. Define $m = \scr Fn_\te{pix}$, the average number of pixels with detectable frequency. If $n+m\geq n_\te{pix}$, then detection is guaranteed. Otherwise, if $\scr M$ Macros impact the moon, the likelihood of any of them being detected is
\begin{align}
\scr P &= 1-\ps{\f{(n_\te{pix}-n)!(n_\te{pix}-m)!}{n_\te{pix}!(n_\te{pix}-n-m)!}}^{\scr M}\,.
\end{align}
\subsection{Homogeneous Limit}
To obtain an estimate of the size of a detectable Macro, we suppose that one ray from each point source on the Macro Trajectory hits a single pixel. In the case that rays don't bend, this is a good approximation for behavior away from the source. Note, however, that this does not account for geometric focusing around the entrance and exit of the Macro trajectory.
\begin{align}
u\leq\sqrt{\f{\Xi E}{\pi^2\rho k v_p^2 R_\te{Moon}^2}}\,.
\end{align}
This expression is independent of the choice of $n_\te{side}$. For the triangle wave, $\sigma_X\leq 4.4\times 10^{-7}\te{\,cm}^2$ is unlikely to be detected unless it enters or exits the Moon adjacent to a seismometer.
\section{Results}
Data has been generated but not run through the analysis routine yet. Preliminary results suggest that the computer analysis will not produce results much stronger than (30).
\section{Conclusion}
For Macros with density sufficient to pass nearly unimpeded through the Moon and sufficiently small cross-section, and in particular nuclear dense objects such as strange quark nuggets with cross-section $\sigma_X\leq 4\times 10^{-7}\te{\, cm}^2$, we have demonstrated that seismological constraints are insufficient to rule them out. Since the Moon (and Earth) preferentially attenuate higher frequency modes, it is unlikely that measuring high frequencies would greatly improve sensitivity to these types of impacts. On the other hand, increasing the lunar seismometer sensitivity to small displacements may allow one to use seismological evidence to rule out high density Macros, although it is uncertain whether seismic noise at such small displacements would make Macro impacts indistinguishable from the background.

>>>>

The moon is an attractive site to carry out a seismological search for Macro impacts. Unlike the Earth, the Moon is nearly dormant. Most internal seismological activity in the Moon originates from deep Moonquakes, which are very attenuated by the time they reach the surface. Most of the noise in lunar seismograms comes from asteroids, but these are for the most part easily identifiable as such. Thus, in the case that we can distinguish between asteroids and Macros, the Moon is a nearly ideal location to place constraints via total seismic event rate. However, in the case that the Macro impacts are very weak, it may be impossible to distinguish them from asteroids or weak Moonquakes based on a single seismogram. In this case, it is necessary to identify correlated events among lunar seismometers which correspond to a propagating source of seismic waves. This sort of analysis will be left to future work.
\section{The Seismic Source}
\subsection{Seismic Wave Generation}
The energy from the initial impact generates a column of melted rock. The super-heated rock, held at constant volume, generates a pressure which sources the seismic waves that can be measured at detectors. Macro's with velocity insignificantly changed by their encounter with the Moon source an overpressure of the form
\begin{align}
\begin{array}{cc}p = p_0 f(r)&\te{ if }a<r<b,-\f h2<z<\f h2\,,\\ p = 0&\te{ else,}\end{array}
\end{align}
where $f$ is an analytic function on $[a,b]$ whose range is a subset of $[-1,1]$, and $p_0$ is the maximum differential pressure in this region. After a short time, the pressure is still of this form, since the inhomogeneity of the Moon is significant only on large scales.

The seismometers on the moon have a net useful range of up to $16\,\te{Hz}$ and are sensitive to displacements as small as $0.3\,\text{nm}$ \cite{latham1973lunar}\cite{nakamura1982apollo}. Therefore, we only calculate the energy contained in the modes of frequency less than $16\,\te{Hz}$.

@@

The choice of $p_0 = 10^8\,\te{Pa}$ does significantly impact our result; ultimately the measured displacements we calculate will be inversely proportional to $p_0$. Our choice of $p_0$, however, is well below the overpressure corresponding to the boundary between linear and non-linear elasticity in the Moon. As the ambient pressure on a sample of rock increases, so too does $p_0$. We have taken $p_0$ to be a factor of $3$ below that of granite with a modest $50\te{ MPa}$ overpressure \cite{lockner200232}. In reality, $p_0$ is probably, on average, several orders of magnitude larger than we claim it to be because the ambient pressure in the Moon is on the order of $\te{GPa}$ rather than $\te{MPa}$, but no measurements, to our knowledge, probe the elastic behavior of rock at these high pressures. It is also important to notice that this choice of $p_0$ corresponds to $r_0$ on the order of $1\te{ km}$, which well exceeds the regime of non-linearity we would have expected based on the solutions to the heat equation. 
\section{Seismic Wave Propagation}
The velocity of $p$-waves as a function of distance from the center $r$ of the Earth and Moon (and presumably other spherical rocky celestial bodies) is of the form $v = a^2 - b^2 r^2$ \cite{garcia2011very}\cite{dziewonski1981preliminary}. Using Snell's law, we obtain the differential equations for the trajectory of $p$-wave rays within such bodies
\begin{align}
v^2 \notag&= \dot r^2 + \f{p_\te{ray}^2 v^4}{r^2}\,,\\
0 &=\dot\theta\pm\f{p_\te{ray} v^2}{r^2}\,.
\end{align}
where $r$ is the radial coordinate of the ray, $\theta$ is the polar angle of the ray measured from the center of the body, and $p_\te{ray}$ is the ray parameter, which is fixed along a ray trajectory. These equations can be integrated. In the case that $v = a^2 - b^2 r^2$, we obtain for some constants $q$ and $\theta_0$.
\begin{align}
\notag&r(t) = \f{a\sqrt{(q^2e^{2abt}-b)^2 + 4a^2b^4 p_\te{ray}}}{b\sqrt{(qe^{2abt}+b)^2 + 4a^2b^4 p_\te{ray}}}\,,\\
&\tan(\theta(t) - \theta_0) = \f{qe^{4abt}-b^2}{4ab^3p_\te{ray}}+abp_\te{ray}\,.
\end{align}
\textit{A priori} these are geodesics on the Poincar\'e disc of radius $\f ab$, which is easily seen from the form of $v$.

@@

\section{Seismic Wave Detection}
Consider a $p$-wave traveling towards positive $x$ and of compact support $S$ in the plane normal to to its motion. Denote $A =\int_S\diff y\diff z$, and let $\chi_S$ be $1$ on $S$ and $0$ elsewhere. The displacement field is
\begin{align}
u(\vec x,t) &= \chi_{\R\times S}\int\f{\diff k}{2\pi}U(k)e^{-\I k(x-v_p t)}\,,\\
\abs{u(\vec x,t)}&\leq\int\f{\diff k}{2\pi}\abs{U(k)}\,,
\end{align}
and the energy is
\begin{align}
E = \rho v_p^2 A\int\f{\diff k}{2\pi}k^2\abs{U(k)}^2\,.
\end{align}
As before, we denote the energy in the low frequency spectrum
\begin{align}
E_k = \rho v_p^2 A\int_0^k\f{\diff k'}{2\pi}k'^2\abs{U(k')}^2\,.
\end{align}
It follows that
\begin{align}
\abs{U(k)} = \sqrt{\f{2\pi}{\rho v_p^2 A}}\f1k\sqrt{\td{E_k}{k}{}}\,,
\end{align}
and from the estimate above
\begin{align}
\abs{u(\vec x,t)}\leq(2\pi\rho v_p^2 A)^{-1/2}\int_0^k\f{\diff k'}{k'}\sqrt{\td{E_{k'}}{k'}{}}\,.
\end{align}
In our case $\Xi E= C_1 k^2$ before attenuation. After anelastic attenuation, $E_k =\te{Ref}\,2C_1 k e^{-k\,\te{Att}}$
\begin{align}
\abs{u(\vec x,t)}\leq\sqrt{\f{2}{\rho v_p^2 A}}\sqrt{\f{\te{Ref}\,\Xi E}{\te{Att}}}\f1k\te{Erf}\ps{\sqrt{\f{\te{Att}\,k}{2}}}\,.
\end{align}
(25) is impossible to integrate when $E_k$ consists of the energies of two different rays, each with different attenuation factors. If attenuation were important, we may average $\te{Att}$ among the coincident rays and obtain an approximate expression. However, since we only wish to consider the lowest frequency modes, $f<16\te{Hz}$, the suppression $\te{Att} k$ is order unity, and does not contribute significantly to $u$. Taking $\te{Att} = 0$, the displacement caused by the incident $p$-wave is bounded above by
\begin{align}
\abs{u(\vec x,t)}&\leq\sqrt{\f{4\te{Ref}\,\Xi E}{\pi k\rho v_p^2 A}}\,.
\end{align}

\section{Simulation}
In previous works, e.g. \cite{banerdt2005seismic}, the homogeneous Earth and Moon models have been used to place constraints of the parameter space of macroscopic dark matter. This approach neglects lensing which occurs because of the velocity gradient and the spherical boundaries of strata, and neglects the energy losses from reflection and refraction across these boundaries. We use the analysis in section III to propagate rays, each carrying a fraction of the total energy deposited by the Macro, to the boundary of the Moon. We then use HEALPix to create an intensity map of the lunar surface for a representative sample of Macro impacts. We then convert the intensity maps to displacement maps, and using the sensitivity of the lunar seismometers, we obtain an average event rate that the lunar seismometers would measure.
\subsection{Data Generation}
We consider a Macro trajectory which passes a distance $D$ from the center of the moon. The Macro trajectory has length $L$, and along its trajectory we consider a set of $M$ points. From each of these points we propagate $N$ randomly oriented rays, each endowed with $E_i = \rho(r_{i})\sigma_Xv_X^2 L/(MN)$ energy, where $i\in\{1,\dots,M\}$ labels the point, and $\rho$ is taken to be a function of the radius. Because of the spherical symmetry, each ray is confined to a subspace of two dimensions. We use the trajectories in section III to propagate the rays. Since the moon is stratified, we must halt the propagation at the boundary of each layer to reflect and transmit rays. During a ray's propagation through a given stratum, its time spent in that layer $\Delta t_i$ is recorded, the attenuation factor in that layer $\te{Att}_i$ is recorded, and the reflection/refraction coefficient $\te{Ref}_i$ is recorded. Each time a ray makes contact with the surface of the Moon, its position, $\te{Att}$, $\te{Ref}$, and $\Delta t$ are recorded. The data produced by the propagation routine is a table of these values and the corresponding surface coordinates, as well as density at the source, $\kappa$ at the source, and $p$-wave velocity at the source. In this model, it takes 16 iterations to propagate a ray from one side of the moon to the other. Rays which propagate from one side of the moon to the other only after 16 iterations will be weak from undergoing 16 refractions. Thus in order to save on computation time, it is reasonable to reduce the number of iterations to 9.
\subsection{Data Analysis}
Analysis begins by converting the position data to HEALPix coordinates for a given $n_\te{side}$. We pick $n_\te{side} = 4$ since this corresponds to an angular resolution of 7 degrees, which accounts for the effects of diffraction about the strata boundaries. Define $A_\te{pix} = 4\pi R_\te{moon}^2/n_\te{pix}$ the area of a HEALPix pixel, where $n_\te{pix} = 12\times 2^{n_{\te{side}}}$. The time it takes a ray to cross a HEALPix pixel can be estimated $t_\te{crossing}^2 = A_\te{pix}/\pi v_\te{surface}^2$. According to VPREMOON, $v_\te{surface} = 1\,\te{km s}^{-1}$. Rays that contact a pixel within $t_\te{crossing}$ of one another are considered to add constructively. Since the integral in section IV is unreasonable to do for more than one ray, we average $\te{Att}$ over the coincident rays. We require that each ray distributes its energy evenly over the area of the HEALPix pixel, and thus set $A = A_\te{pix}$. This is a good approximation when $N=n_\te{pix}$, since the average angular separation of the rays at the source corresponds to the average angular separation of the HEALPix pixels.

@@

This procedure is repeated for 17 values of $D$ spaced evenly by $100\,\te{km}$. The average fraction of detectable surface area is then approximately
\begin{align}
\scr F =\f{\sum_D \scr F_D}{\sum_D D}\,.
\end{align}
If there are $n$ lunar seismometers, we suppose that each occupies its own HEALPix pixel. Define $m = \scr Fn_\te{pix}$, the average number of pixels with detectable frequency. If $n+m\geq n_\te{pix}$, then detection is guaranteed. Otherwise, if $\scr M$ Macros impact the moon, the likelihood of any of them being detected is
\begin{align}
\scr P &= 1-\ps{\f{(n_\te{pix}-n)!(n_\te{pix}-m)!}{n_\te{pix}!(n_\te{pix}-n-m)!}}^{\scr M}\,.
\end{align}
\subsection{Homogeneous Limit}
To obtain an estimate of the size of a detectable Macro, we suppose that one ray from each point source on the Macro Trajectory hits a single pixel. In the case that rays don't bend, this is a good approximation for behavior away from the source. Note, however, that this does not account for geometric focusing around the entrance and exit of the Macro trajectory.
\begin{align}
u\leq\sqrt{\f{\Xi E}{\pi^2\rho k v_p^2 R_\te{Moon}^2}}\,.
\end{align}
This expression is independent of the choice of $n_\te{side}$. For the triangle wave, $\sigma_X\leq 1\times 10^{-7}\te{\,cm}^2$ is unlikely to be detected unless it enters or exits the Moon adjacent to a seismometer.
\section{Results}
Data has been generated but not run through the analysis routine yet. Preliminary results suggest that the computer analysis will not produce results much stronger than (30).
\section{Conclusion}
For Macros with density sufficient to pass nearly unimpeded through the Moon and sufficiently small cross-section, and in particular nuclear dense objects such as strange quark nuggets with cross-section $\sigma_X\leq 1\times 10^{-7}\te{\, cm}^2$, we have demonstrated that seismological constraints are insufficient to rule them out. Since the Moon (and Earth) preferentially attenuate higher frequency modes, it is unlikely that measuring high frequencies would greatly improve sensitivity to these types of impacts. On the other hand, increasing the lunar seismometer sensitivity to small displacements may allow one to use seismological evidence to rule out high density Macros, although it is uncertain whether seismic noise at such small displacements would make Macro impacts indistinguishable from the background.
